---
title: "Lab 17"
author: "Bomin Xie"
format: pdf
---

# Getting started

```{r}
# Import vaccination data
vax <- read.csv("https://marcos-diazg.github.io/BIMM143_SP23/class-material/class17/covid19vaccinesbyzipcode_test.csv")
head(vax)
```

Q1: The column details the total number of people fully vaccinated are "persons_fully_vaccinatted".

Q2: The column details the Zip code tabulation area is "zip_code_tabulation_area".

Q3: The earliest date in this dataset is "2021-01-05".

Q4: The latest date in this dataset is "2023-05-23".

```{r}
# install.packages("skimr")
library(skimr)
skimr::skim_without_charts(vax)
```

Q5: In this dataset, there are 14 numeric columns in this dataset.

Q6: There are 17711 "missing values" in the "persons_fully_vaccinated" column.

Q7: Based on the skimr result, there are 8.04 percent of  "persons_fully_vaccinated" values are missing.

## Working with dates

```{r}
# install.packages("lubridate")
library(lubridate)
today()
vax$as_of_date <- ymd(vax$as_of_date)
today() - vax$as_of_date[nrow(vax)]
length(unique(vax$as_of_date))
```

Q9: It has been 21 days passed since the last update.

Q10: There are 125 unique dates in the dataset.

# Working with ZIP codes

```{r}
# install.packages("zipcodeR")
library(zipcodeR)
geocode_zip('92037')
zip_distance('92037','92109')
```

# Focus on San Diego area

```{r}
library(dplyr)
sd <- filter(vax, county == "San Diego")
nrow(sd)
```
```{r}
sd.10 <- filter(vax, county == "San Diego" &
                age5_plus_population > 10000)
```

```{r}
length(unique(sd$zip_code_tabulation_area))
sd$zip_code_tabulation_area[which.max(unique(sd$tot_population))]
```

Q11: There are 107 distinct zip codes.

Q12: The zip code area with largest population in this dataset is 92154.

```{r}
avg_percent <- filter(sd, as_of_date == "2023-02-28")
mean(avg_percent$percent_of_population_fully_vaccinated, na.rm = TRUE) * 100
```


Q13: The overall average "percent of population fully vaccinated value" is 74.13.

Q14: 

```{r}
library(ggplot2)
ggplot(avg_percent, aes(percent_of_population_fully_vaccinated)) + 
  geom_histogram(bins = 10, na.rm = TRUE, color= "black", fill = "grey") + 
  ggtitle("Histogram of Vaccination rates across San Diego County") + 
  xlab("Percent of Population Fully Vaccinated in a Zip Code Area") + ylab("Count(Zip Code Areas")
```

## Focus on UCSD/La Jolla

```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```

Q15:

```{r}
vaccination_rate_plot <- ggplot(ucsd) +
  aes(as_of_date, percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(title = "Vaccination Rate for La Jolla CA 92037", x= "Date", y="Percent Vaccinated")
vaccination_rate_plot
```

## Comparing to similar sized areas

```{r}
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2023-02-28")

#head(vax.36)
```

Q16:

```{r}
mean_92037 <- mean(vax.36$percent_of_population_fully_vaccinated)
mean_92037
vaccination_rate_plot + geom_hline(yintercept = mean_92037, color = "red", linetype = 5)
```

Q17: The 6 number summary is listed below:

```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```

Q18:

```{r}
ggplot(vax.36, aes(percent_of_population_fully_vaccinated)) +
  geom_histogram(na.rm = TRUE) + 
  xlim(0,1) + 
  xlab("Percent Vaccinated") + ylab("Count")
        
```

```{r}
vax %>% filter(as_of_date == "2023-05-23") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)

vax %>% filter(as_of_date == "2023-05-23") %>%  
  filter(zip_code_tabulation_area=="92109") %>%
  select(percent_of_population_fully_vaccinated)
```

Q19: Based on the above result, both the two ZIP code areas are below the average value of 0.7219.

Q20: 

```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)


ggplot(vax.36.all) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="Blue", na.rm = TRUE) +
  ylim(0,1) +
  labs(x="Date", y="Percent Vaccinated",
       title="Vaccination Rate Across California",
       subtitle="Only areas with population above 36k are shown") +
  geom_hline(yintercept = mean_92037, linetype=5)
```

