---
title: "Lab 15"
author: "Bomin Xie A16144147"
format: pdf
---

# UNIX for bioinformatics

1. The star character "*" is used as a wildcard character to represent any sequence of characters. Right here, it is used to match and expand all the files in the current directory that have a ".faa.gz" extension.

2. There are 41 sequences in ```mouse.1.protein.faa```.

3. Running without ```> mm-first.fa``` part, it will print out the first 11 line of the file without saving it.

4. It will save the first 11 line of the file into ```mm-first.fa``` file.

5. I use the command ```grep -c . mm-second.fa``` to determine the number of sequences.

```{r}
blast_zebrafish <- read.csv("mm-second.x.zebrafish.tsv", sep="\t")
colnames(blast_zebrafish) = c("qseqid", "sseqid", "pident", "length", "mismatch", "gapopen", "qstart", "qend", "sstart", "send", "evalue", "bitscore")
```


```{r}
library(ggplot2)
ggplot(blast_zebrafish, aes(pident, bitscore)) + geom_point(alpha=0.1) 
ggplot(blast_zebrafish, aes((blast_zebrafish$pident * (blast_zebrafish$qend - blast_zebrafish$qstart)), bitscore)) +
geom_point(alpha=0.1) + geom_smooth()
```

6. The -r option stands for "recursive," that allow scp to copy directories and their contents recursively, ensuring that all files and subdirectories within the specified directory are copied to the destination. The * character is used as a wildcard to represent any sequence of characters in the filename, which allows the match of multiple files or directories.