---
title: "Homework 11"
author: "Bomin Xie A16144147"
format: pdf
---

# Homework 11

You are interested in assessing genetic differences on a population scale. So, you processed about ~230 samples and did the normalization on a genome level. Now, you want to find whether there is any association of the 4 asthma-associated SNPs (rs8067378...) on ORMDL3 expression.

```{r}
expression = read.table("https://marcos-diazg.github.io/BIMM143_SP23/class-material/class11/rs8067378_ENSG00000172057.6.txt")
summary(expression)
median(expression$exp[expression$geno == "A/A"])
median(expression$exp[expression$geno == "A/G"])
median(expression$exp[expression$geno == "G/G"])
table(expression$geno)
```

Q13: The sample size for each genotype is: 108 samples for the A/A genotype, 233 for A/G, and 121 for G/G. The median of each genotype is 31.2 (A/A), 25.1 (A/G), and 20.1 (G/G).

```{r}
library(ggplot2)
ggplot(expression) + aes(geno, exp, fill=geno) + 
  geom_boxplot(notch = TRUE) +
  geom_jitter(color="black", alpha=0.2, width = 0.2) +
  xlab("Genotype") + ylab("Expression") + labs(fill = "Genotype")
```

Q14: From the graph, A/A has higher expression on ORMDL3 than G/G genotype. The genotype difference is correlated to the expression differences, which indicates the SNP might affect the ORMDL3 expression.

