---
title: "Lab05"
author: "Bomin Xie"
format: pdf
---

# Week 4 Lab 5

Initiation of ggplot2, and visit the data set:

```{r}
#install.pockages("ggplot2")
library(ggplot2)
#View(cars)
plot(cars) # A quick R-based plot
ggplot(cars)
```

Plot dataset with ggplot2:

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point()
```

Plot data points onto ggplot2 with aes() and geom_point():

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth() # Add line
```

```{r}
p <- ggplot(cars) +
      aes(x=speed, y=dist) +
      geom_point() +
      geom_smooth(method = "lm", se=FALSE) 
```

```{r}
p <- ggplot(cars) +
      aes(x=speed, y=dist) +
      geom_point() +
      geom_smooth(method = "lm", se=FALSE) +
      labs(title="Speed and Stopping Distances of Cars",
          x="Speed (MPH)", 
          y="Stopping Distance (ft)")
```

------------------------------------------------------------------------

Read the drug expression data:

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

Analyzing the dataset:

```{r}
nrow(genes) # Number of rows in the data set
colnames(genes) # Name of columns in the data set
table(genes$State) # Different states of genes in the data set
table(genes$State) / nrow(genes) * 100
```

Plot a scatter plot of the two conditions in the genes data

```{r}
ggplot(genes) + 
    aes(x=Condition1, y=Condition2, col=State) +
    geom_point() +
    scale_colour_manual(values = c("blue","grey","red")) +
    labs(title="Gene Expresion Changes Upon Drug Treatment",
         x="Control (no drug) ",
         y="Drug Treatment")
```

---

Advanced ggplot() function through gapminder dataset

```{r}
# Import gapminder
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"
gapminder <- read.delim(url)
```

Use dplyr to focus on single year:

```{r}
# install.packages("dplyr") 
library(dplyr)
gapminder_2007 <- gapminder %>% filter(year==2007)
```

Plot the result:

```{r}
ggplot(gapminder_2007) +
  aes(x=gdpPercap, y=lifeExp) +
  geom_point()
```

Adding more aesthetics:

```{r}
ggplot(gapminder_2007) +
  aes(x=gdpPercap, y=lifeExp, color=continent, size=pop) +
  geom_point(alpha=0.5)
```

---

Analyzing data through bar chart:

```{r}
gapminder_top5 <- gapminder %>% 
  filter(year==2007) %>% 
  arrange(desc(pop)) %>% 
  top_n(5, pop)

gapminder_top5
ggplot(gapminder_top5) + 
  geom_col(aes(x = country, y = pop, fill = continent))
```

---

Animations of ggplot, through gifski() and gganimate():

```{r}
library(gifski)
library(gganimate)

ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, colour = country)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
# scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  facet_wrap(~continent) +
  labs(title = 'Year: {frame_time}', x = 'GDP per capita', y = 'life expectancy') +
  transition_time(year) +
  shadow_wake(wake_length = 0.1, alpha = FALSE)
```

